---
title: "Hantavirus Data Analysis"
author: "Andrew Ignatescu"
format: html
editor: visual
---

## Introduction

Hantavirus is a rare but deadly virus transmitted to humans through contact with infected rodents. Since its first major outbreak in 1993 in the Four Corners region of the American Southwest, the CDC has tracked cases across the United States. This research explores where and when Hantavirus outbreaks have occurred, using CDC data spanning from before 1993 to 2023

## About the Data

This analysis uses Hantavirus case data published by the Centers for Disease Control and Prevention (CDC). The data was downloaded from:

<https://www.cdc.gov/hantavirus/data-research/cases/index.html>

The dataset contains individual case records from before 1993 through 2023, including illness onset date, state FIPS code, and patient outcome.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(stringr)
library(maps)
library(lubridate)
library(forecast)
library(knitr)

hantavirus_data <- read.csv("data/HantaVirusTableData.csv", skip = 1)

str(hantavirus_data)
head(hantavirus_data)
dim(hantavirus_data)
```

## Cases Per Year

Hantavirus is a rare but deadly disease transmitted through contact with infected rodents. This chart shows the number of confirmed cases reported in the United States from before 1993 to 2023. The spike in 1993 marks the first major outbreak in the Four Corners region of the American Southwest.

```{r}
#| message: false
#| warning: false
ggplot(hantavirus_data, aes(x = factor(
  ifelse(grepl("Before", IllnessOnsetDate),
         "Before 1993",
         str_extract(IllnessOnsetDate, "\\d{4}")),
  levels = c("Before 1993", as.character(1993:2024))))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Hantavirus Cases by Year From Before 1993 to 2023",
       x = "Year", y = "Number of Cases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Cases Over Time

This line chart shows the trend in Hantavirus cases by year, excluding pre-1993 cases which lack specific dates.

```{r}
#| message: false
#| warning: false
cases_by_year <- hantavirus_data %>%
  filter(!grepl("Before", IllnessOnsetDate), !is.na(IllnessOnsetDate)) %>%
  mutate(Year = as.integer(str_extract(IllnessOnsetDate, "\\d{4}"))) %>%
  filter(!is.na(Year)) %>%
  count(Year, name = "Cases")

ggplot(cases_by_year, aes(x = Year, y = Cases)) +
  geom_line(color = "red", linewidth = 1.2) +
  geom_point(color = "red", size = 3) +
  scale_x_continuous(breaks = seq(min(cases_by_year$Year),
                                   max(cases_by_year$Year), by = 2)) +
  labs(title = "Hantavirus Cases Over Time (1993-2023)",
       subtitle = "Pre-1993 cases excluded due to undated records",
       x = "Year", y = "Number of Cases") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

## Cases Per State

This map shows the geographic distribution of Hantavirus cases across the United States.

```{r}
#| message: false
#| warning: false
state_fips <- data.frame(
  StateFIPS = c("04","06","08","12","16","17","18","19","20",
                "22","25","26","27","30","31","32","35","36",
                "37","38","40","41","42","46","48","49","50",
                "53","55","56"),
  state = c("arizona","california","colorado","florida","idaho",
            "illinois","indiana","iowa","kansas","louisiana",
            "massachusetts","michigan","minnesota","montana",
            "nebraska","nevada","new mexico","new york",
            "north carolina","north dakota","oklahoma","oregon",
            "pennsylvania","south dakota","texas","utah","vermont",
            "washington","wisconsin","wyoming")
)

state_counts <- hantavirus_data %>%
  mutate(StateFIPS = sprintf("%02d", as.integer(StateFIPS))) %>%
  count(StateFIPS) %>%
  left_join(state_fips, by = "StateFIPS")

us_map <- map_data("state")
map_merged <- left_join(us_map, state_counts, by = c("region" = "state"))

ggplot(map_merged, aes(x = long, y = lat, group = group, fill = n)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkred",
                      na.value = "grey90", name = "Cases") +
  labs(title = "Hantavirus Cases by State (1993-2023)") +
  theme_void()
```

## Hantavirus Cases Per Million by State

```{r}
#| message: false
#| warning: false
state_populations <- data.frame(
  state = c("arizona","california","colorado","florida","idaho",
            "illinois","indiana","iowa","kansas","louisiana",
            "massachusetts","michigan","minnesota","montana",
            "nebraska","nevada","new mexico","new york",
            "north carolina","north dakota","oklahoma","oregon",
            "pennsylvania","south dakota","texas","utah","vermont",
            "washington","wisconsin","wyoming"),
  population = c(7151502, 39538223, 5773714, 21538187, 1839106,
                 12812508, 6785528, 3190369, 2937880, 4657757,
                 7029917, 10077331, 5706494, 1084225, 1961504,
                 3104614, 2117522, 20201249, 10439388, 779094,
                 3959353, 4237256, 13002700, 886667, 29145505,
                 3271616, 643077, 7705281, 5893718, 576851)
)

cases_per_million <- state_counts %>%
  left_join(state_populations, by = "state") %>%
  mutate(per_million = (n / population) * 1000000)

map_per_million <- left_join(us_map, cases_per_million, by = c("region" = "state"))

ggplot(map_per_million, aes(x = long, y = lat, group = group, fill = per_million)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkred",
                      na.value = "grey90", name = "Cases per Million") +
  labs(title = "Hantavirus Cases per Million by State (1993-2023)") +
  theme_void()
```

## Hantavirus Cases Over Time by Top 5 States

```{r}
#| message: false
#| warning: false
top5_states <- state_counts %>%
  slice_max(n, n = 5) %>%
  pull(state)

top5_data <- hantavirus_data %>%
  mutate(StateFIPS = sprintf("%02d", as.integer(StateFIPS))) %>%
  left_join(state_fips, by = "StateFIPS") %>%
  filter(state %in% top5_states) %>%
  filter(!grepl("Before", IllnessOnsetDate), !is.na(IllnessOnsetDate)) %>%
  mutate(Year = as.integer(str_extract(IllnessOnsetDate, "\\d{4}"))) %>%
  filter(!is.na(Year)) %>%
  count(Year, state, name = "Cases")

ggplot(top5_data, aes(x = Year, y = Cases, color = state, group = state)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(1993, 2023, by = 2)) +
  labs(title = "Hantavirus Cases Over Time for Top 5 States",
       x = "Year", y = "Number of Cases", color = "State") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Hantavirus Case Predictions 2024-2030

```{r}
#| message: false
#| warning: false
ts_data <- ts(cases_by_year$Cases, start = min(cases_by_year$Year))
fit <- auto.arima(ts_data)
future <- forecast(fit, h = 7)

forecast_df <- data.frame(
  Year = 2024:2030,
  Cases = as.numeric(future$mean),
  Lower = as.numeric(future$lower[, 2]),
  Upper = as.numeric(future$upper[, 2]),
  Type = "Forecast"
)

historical_df <- cases_by_year %>%
  mutate(Type = "Historical")

ggplot() +
  geom_line(data = historical_df,
            aes(x = Year, y = Cases),
            color = "steelblue", linewidth = 1) +
  geom_line(data = forecast_df,
            aes(x = Year, y = Cases),
            color = "red", linewidth = 1, linetype = "dashed") +
  geom_ribbon(data = forecast_df,
              aes(x = Year, ymin = Lower, ymax = Upper),
              fill = "red", alpha = 0.2) +
  scale_x_continuous(breaks = seq(1993, 2030, by = 2)) +
  labs(title = "Hantavirus Cases with Forecast 2024-2030",
       subtitle = "Shaded area shows 95% prediction interval",
       x = "Year", y = "Number of Cases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Top 10 States by Case Count

```{r}
#| message: false
#| warning: false
top10_states <- hantavirus_data %>%
  mutate(StateFIPS = sprintf("%02d", as.integer(StateFIPS))) %>%
  left_join(state_fips, by = "StateFIPS") %>%
  filter(!is.na(state)) %>%
  count(state, name = "Cases") %>%
  arrange(desc(Cases)) %>%
  slice(1:10) %>%
  mutate(state = str_to_title(state)) %>%
  rename(State = state)

kable(top10_states, caption = "Top 10 States by Hantavirus Case Count (1993-2023)")
```
